Modeling Ionic Polymer-Metal Composites\\ with Space-Time Adaptive Multimesh \emph{hp}-FEM

### 2011-02-16 response.tex

\documentclass[mathpazo]{cicp}

%\journal{Applied Mathematics and Computation}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsrefs}
%\usepackage{subfig}

\begin{document}

\title{Response to the reviewers of \emph{Modeling Ionic Polymer-Metal Composites\\ with Space-Time Adaptive Multimesh hp-FEM}}

\author[Deivid Pugal et.~al.]{Deivid Pugal\affil{1}\comma\affil{4},
Pavel Solin\affil{2}\comma\affil{3}\comma\corrauth,
Kwang J. Kim\affil{1}, and Alvo Aabloo\affil{4}}

\address{\affilnum{1}\ Mechanical Engineering Department, University of Nevada, Reno, NV, U.S.A.\\
\affilnum{2}\ Department of Mathematics and Statistics, University of Nevada, Reno, NV, U.S.A.\\
\affilnum{3}\ Institute of Thermomechanics, Prague, Czech Republic\\
\affilnum{4}\ Institute of Technology, Tartu University, Estonia}

\emails{{\tt david.pugal@gmail.com} (D.~Pugal), {\tt solin@unr.edu} (P.~Solin),
	{\tt kwangkim@unr.edu} (K.~Kim), {\tt alvo@ut.ee} (A.~Aabloo)}

\begin{abstract}
We are concerned with a model of ionic polymer-metal composite (IPMC) materials
that consists of a coupled system of the Poisson and Nernst-Planck equations, 
discretized by means of the finite element method (FEM). We show that due to the 
transient character of the problem it is efficient to use adaptive algorithms 
that are capable of changing the mesh dynamically in time. We also show 
that due to large qualitative and quantitative differences between the 
two solution components, it is efficient to approximate them on different 
meshes using a novel adaptive multimesh \emph{hp}-FEM. The study is 
accompanied with numerous computations and comparisons of the adaptive 
multimesh \emph{hp}-FEM with several other adaptive FEM algorithms. 
\end{abstract}

\maketitle

\section{Reviewer \#1}

\begin{enumerate}
\item \emph{The comparisons presented by the authors are interesting. One important aspect regarding the
Poisson-Nernst-Planck equations might be missing. A characteristic quantity of the system is the
dimensionless Debye length. It is well-known that the case of small Debye length’s provides severe
numerical difficulties. Hence, it would be very interesting to see how hp-FEM are able to handle
such cases. This might also lead to a different picture where possibly the electrostatic potential
needs more adaptivity than the concentration of counter ions in contrast to examples considered so
far.}

TODO
\item \emph{last sentence and Figure 1 Could the authors give a schematic picture presenting the electrostatic forces causing the deformation/bending of the IPMC?}

An explanation was added to the images in Figure 1.

\item \emph{3rd line "materials that includes"}

Fixed.

\item \emph{2nd paragraph, The computing power ... Are there references which elucidate this statement?}

A reference was added to a previous study where solving the PNP system in 3D was discussed.

\item \emph{1st sentence, ... principal difficulties ... As mentioned above, there are also other issues
known for the Poisson-Nernst-Planck system in the case of small Debye lenghts. Are the
authors sure that the motivation given in Section is numerical more demanding? If yes, are
there references available?}

TODO

\item \emph{eq. (3.5) It might be more appropriate to present the dimensionless formulation of the problem instead of introducing the physically less motivated notations (3.5). Such a formulation then
immediately provides the characteristic dimensionless parameter of the system, i.e. the Debye
length.}

This was very good point. The dimensionless form of the equations was derived. $\mathbf{All}$ the
results were recalculated and each graph and image in the Results section was redone.
Instead of using $C$
and $\phi$, the corresponding scaled variables $c$ and $\varphi$ are now used throughout the results section.

Furthermore, this time, all the results were calculated using adaptivity where the relative error
of the physical fields was considered separately.  Previously, absolute errors were compared
and the absolute error of the physical field $\phi$ was always very  much smaller compared to
$C$. Therefore no adaptivity of whatsoever was observed in case of $\phi$. The new results
(for instance, Figures~14 and~15) show that the mesh and the polynomial
space of physical field $\varphi$ is also adapted, however, not as much as in case of $c$.
We feel that this approach makes physically more sense.

\item \emph{3rd paragraph The reader might loose at this point the interest in reading further since there
is no motivation which favors a hp-FEM discretization over traditional low-order FEM. One
sentence regarding computational performance or a related comment might be useful ...}

TODO Pavel

\item \emph{2nd last sentence in 1st paragraph A short explanation, how Hermes further deals with this
difference without going to much into details, might be interesting here ...}

TODO Pavel

\item \emph{1st sentence in Sec. 5.2 Convergence plots of the error, which the authors mention, would be interesting...}

Figure 13 was added to show an example error convergence in case of HP\_ANISO.

\item \emph{last paragraph "The results"}

Fixed.

\item \emph{Section 5.4, eq. (5.4) What does Hermes do in the case $e^n\geq\delta$?}

Currently Hermes does not support the time step reduction. However, this is planned to be added
in future. A comment was also added below (5.5): \emph{At this point,
the implementation does not support reducing the time step when $e^n \geq \delta$.}

\end{enumerate}

\section{Reviewer \#2}

\begin{enumerate}

\item \emph{The authors have not given the initial values of $C$}

The initial value was added below Eq.~(1.1).

\item \emph{The authors have not given the size of time-step and the relation
of time step size and h; p or CFL stability condition.}

TODO Pavel?

\end{enumerate}
\end{document}

### 2011-02-16 intro.tex
Half of the graphs redone in nondimensional form

The Nernst-Planck equation for a mobile species ---
in our case for counter ions --- has the form
\begin{equation}
  \frac{\partial C}{\partial t}+\nabla\cdot(-D\nabla C-\mu FC\nabla\phi)=0.
  \label{eq:nernst-planck}
\end{equation}
Here $C$ stands for the counter ion concentration, $D$ is diffusion, $\mu$ mobility,
$F$ Faraday constant, and $\phi$ voltage. We have neglected 
the velocity of the species as in our case it can be assumed zero. 
The Poisson equation has the form
\begin{equation}
  -\nabla^2\phi=\frac{F\rho}{\varepsilon}
  \label{eq:poisson}
\end{equation}
where $\varepsilon$ is the absolute dielectric permittivity. The
charge density $\rho = C-C_{0}$ 
where $C_{0}$ is a constant anion concentration.

>>>>

The Nernst-Planck equation for a mobile species ---
in our case for counter ions --- has the form
\begin{equation}
  \frac{\partial C}{\partial t}+\nabla\cdot(-D\nabla C-\mu FC\nabla\phi)=0.
  \label{eq:nernst-planck}
\end{equation}
Here $C$ stands for the counter ion concentration with the initial value of $C_0$,
$D$ is diffusion, $\mu$ mobility,
$F$ Faraday constant, and $\phi$ voltage. We have neglected 
the velocity of the species as in our case it can be assumed zero. 
The Poisson equation has the form
\begin{equation}
  -\nabla^2\phi=\frac{F\rho}{\varepsilon}
  \label{eq:poisson}
\end{equation}
where $\varepsilon$ is the absolute dielectric permittivity. The
charge density $\rho = C-C_{0}$ 
where $C_{0}$ is a constant anion concentration.

### 2011-02-16 motivation.tex
Half of the graphs redone in nondimensional form

\begin{table}[!ht]
\caption{Constants used in the Poisson-Nernst-Planck system of equations.}
\centering
\label{Table:used-constants}
{
\begin{tabular}{llll}
  \hline \hline
  Constant&Value&Unit&Description\\
  \hline
  $D$&$10\times10^{-11}$&$\frac{m^2}{s}$&Diffusion constant\\
  $z$&1&-&Charge number\\
  $F$&96,485&$\frac{C}{mol}$&Faraday number\\
  $R$&8.31&$\frac{J}{mol\cdot K}$&The gas constant\\
  $\mu\ \left( = \frac{D}{RT}\right)$&$4.11\times 10^{-14}$&$\frac{s}{mol\cdot K}$&Mobility\\
  $C_{0}$&1,200&$\frac{mol}{m^3}$&Anion concentration\\
  $\varepsilon$&0.025&$\frac{F}{m}$&Electric permittivity\\
  \hline
  \hline
\end{tabular}
}
\end{table}

>>>>

\begin{table}[!ht]
\caption{Constants used in the Poisson-Nernst-Planck system of equations.}
\centering
\label{Table:used-constants}
{
\begin{tabular}{llll}
  \hline \hline
  Constant&Value&Unit&Description\\
  \hline
  $D$&$10\times10^{-11}$&$\frac{m^2}{s}$&Diffusion constant\\
  $z$&1&-&Charge number\\
  $F$&96,485&$\frac{C}{mol}$&Faraday number\\
  $R$&8.31&$\frac{J}{mol\cdot K}$&The gas constant\\
  $\mu\ \left( = \frac{D}{RT}\right)$&$4.11\times 10^{-14}$&$\frac{s}{mol\cdot K}$&Mobility\\
  $C_{0}$&1,200&$\frac{mol}{m^3}$&Anion concentration\\
  $\varepsilon$&0.025&$\frac{F}{m}$&Electric permittivity\\
  $l$&$200\times10^{-6}$&$m$& length scale\\
  \hline
  \hline
\end{tabular}
}
\end{table}

### 2011-02-16 results.tex
Half of the graphs redone in nondimensional form

The solutions to the PNP problem exhibit a specific behavior that was 
described above. In order to find the best adaptive method to deal with 
this type of problems, we performed numerous computations using all 
adaptivity modes in both the single-mesh and multi-mesh regimes.
In the numerical experiments we paid attention to the 
relative error, cumulative CPU time, and problem size 
in terms of number of degrees of freedom (DOF) in each 
time step. The dimensional variables $x$, $y$, $t$, $C$,
and $\phi$ are used to present the solutions.

@@

An example of the solution at $t=0.7\ s$ and $t=3.0\ s$ 
calculated with the HP\_ANISO refinement mode is shown
in Figs.~\ref{fig:cphi-1}~and~\ref{fig:cphi-2}.

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{cphi-1}
  \caption{\label{fig:cphi-1} Concentration $C$
  and voltage $\phi$ at $t=0.7\ s$.}
  \end{centering}
\end{figure}

@@

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{cphi-2}
  \caption{\label{fig:cphi-2} Concentration $C$
  and voltage $\phi$ at $t=3.0\ s$.}
  \end{centering}
\end{figure}

The reader can see that at $t=0.7\ s$ some ionic migration has already 
taken place and large concentration gradients near the boundaries $\partial\Omega_1$ 
and $\partial\Omega_3$ have formed. The figures also show that the meshes 
at $t=0.7\ s$ and $t=3.0\ s$ are different.

@@

\noindent
It can be seen that $hp$-FEM results in a shorter computing 
time and smaller number
of DOF than the low-order FEM. The same holds true
for H\_ISO and HP\_ISO modes. In fact, in case of H\_ISO the relative error did not
converge to the pre-set threshold value of $0.5\%$.
Therefore, the $h$-FEM solutions will be omitted from the further comparisons.
Instead, only $hp$-FEM solutions on the coarse mesh and $p$-FEM solutions
on the fine mesh will be discussed. It must be also noted that the error
converged to or below $0.5\%$ for all $p$-FEM and $hp$-FEM results, therefore
the graphs of the error as a function of physical time will not be presented.

\subsection{Comparison of single-mesh and multi-mesh $hp$-FEM}
% David: Error convergence plots here.
Running the simulation with different adaptivity modes 
and meshes showed that the multi-mesh $hp$-FEM configuration resulted in
the smallest problems, shortest computing times, and better or the same error 
convergence compared to any single-mesh configuration.
This is illustrated for HP\_ANISO adaptivity mode in Fig.~\ref{fig:singlemultidof}, Fig.~\ref{fig:singlemulticpu} and Fig.~\ref{fig:singlemultierror}.
The same holds true for HP\_ISO mode.

@@

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{singlemulti_error}
  \caption{\label{fig:singlemultierror} Relative solution error as a function 
  of physical time for single-mesh and multi-mesh configurations with 
  HP\_ANISO adaptivity mode.}
  \end{centering}
\end{figure}
\noindent 
Fig. \ref{fig:poly} shows higher-order meshes in the adaptive multi-mesh $hp$-FEM
computation for $C$ and $\phi$ at $t = 0.7$ s. Different 
colors mean different polynomial degrees. A diagonal pattern inside an element 
tells that the element has different polynomial degrees in the 
horizontal and vertical directions.

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=.75\columnwidth]{poly}
  \caption{\label{fig:poly} Higher-order FEM mesh for 
  $C$ and $\phi$ at $t=0.7\ s$. }
  \end{centering}
\end{figure}

The result are in good agreement with Fig.~\ref{fig:cphi-2} --- in the vicinity
of the boundaries $\partial \Omega_1$ and $\partial\Omega_3$, the concentration gradient
is much greater than the voltage gradient. Therefore the multi-mesh $hp$-FEM adaptivity 
algorithm has increased the maximum polynomial degree for the $C$-space to~7 while 
the maximum polynomial degree for the $\phi$-space is~2. One can also see that the mesh 
is significantly more refined for $C$.
Since these results are representative for all adaptivity modes, only multi-mesh 
configurations are considered in the following.

>>>>

The solutions to the PNP problem exhibit a specific behavior that was 
described above. In order to find the best adaptive method to deal with 
this type of problems, we performed numerous computations using all 
adaptivity modes in both the single-mesh and multi-mesh regimes.
In the numerical experiments we paid attention to the 
relative error, cumulative CPU time, and problem size 
in terms of number of degrees of freedom (DOF) in each 
time step. The scaled variables $c$
and $\varphi$ and the unscaled time $t$ are used to present the solutions.

@@

An example of the solution at $t=0.1\ s$ and $t=3.0\ s$ 
calculated with the HP\_ANISO refinement mode is shown
in Figs.~\ref{fig:cphi-1}~and~\ref{fig:cphi-2}.

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{cphi-1}
  \caption{\label{fig:cphi-1} Scaled concentration $c$
  and voltage $\varphi$ at $t=0.1\ s$.}
  \end{centering}
\end{figure}

@@

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{cphi-2}
  \caption{\label{fig:cphi-2} Scaled concentration $c$
  and voltage $\varphi$ at $t=3.0\ s$.}
  \end{centering}
\end{figure}

The reader can see that at $t=0.1\ s$ some ionic migration has already 
taken place and large concentration gradients near the boundaries $\partial\Omega_1$ 
and $\partial\Omega_3$ have formed. The figures also show that the meshes 
at $t=0.1\ s$ and $t=3.0\ s$ are different.

@@

\noindent
It can be seen that $hp$-FEM results in a shorter computing 
time and smaller number
of DOF than the low-order FEM. The same holds true
for H\_ISO and HP\_ISO modes. In fact, in case of H\_ISO the relative error did not
converge to the pre-set threshold value of $0.5\%$.
Therefore, the $h$-FEM solutions will be omitted from the further comparisons.
Instead, only $hp$-FEM solutions on the coarse mesh and $p$-FEM solutions
on the fine mesh will be discussed.

\subsection{Comparison of single-mesh and multi-mesh $hp$-FEM}
% David: Error convergence plots here.
Running the simulation with different adaptivity modes 
and meshes showed that the multi-mesh $hp$-FEM configuration resulted in
the smallest problems and similar error 
convergence compared to any single-mesh configuration. However,
multi-mesh problems generally resulted in longer computing times. This is a known
shortcoming of Hermes at this point and it is due to the fact that
multi-mesh uses the union mesh (see Section~\ref{sec:hermes}) where 
the numerical integration of high order is done on very small elements.
The problem size and computing time are illustrated for HP\_ANISO adaptivity mode in Fig.~\ref{fig:singlemultidof} and Fig.~\ref{fig:singlemulticpu}.
The same holds true for HP\_ISO mode.  It must be also noted that the error
converged to or below $0.5\%$ for all $p$-FEM and $hp$-FEM results. This is illustrated in
Fig.~\ref{fig:singlemultierror} for $hp$-FEM single and multi-mesh configurations.

@@

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{singlemulti_error}
  \caption{\label{fig:singlemultierror} Relative solution error as a function 
  of physical time for single-mesh and multi-mesh configurations with 
  HP\_ANISO adaptivity mode.}
  \end{centering}
\end{figure}
\noindent 
Figs.~\ref{fig:poly} and~\ref{fig:poly2} show higher-order meshes in the adaptive multi-mesh $hp$-FEM
computation for $c$ and $\varphi$ at $t = 0.1$~s and $t=3.0$~s, respectively. Different 
colors mean different polynomial degrees. A diagonal pattern inside an element 
tells that the element has different polynomial degrees in the 
horizontal and vertical directions.

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=.75\columnwidth]{poly}
  \caption{\label{fig:poly} Higher-order FEM mesh for 
  $c$ and $\varphi$ at $t=0.1\ s$. }
  \end{centering}
\end{figure}

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=.75\columnwidth]{poly2}
  \caption{\label{fig:poly2} Higher-order FEM mesh for 
  $c$ and $\varphi$ at $t=3.0\ s$. }
  \end{centering}
\end{figure}
\noindent
The result are in good agreement with Fig.~\ref{fig:cphi-2} --- in the vicinity
of the boundaries $\partial \Omega_1$ and $\partial\Omega_3$, the concentration gradient
is much greater than the voltage gradient. Therefore at $t=0.1$~s, 
the multi-mesh $hp$-FEM adaptivity 
algorithm has increased the maximum polynomial degree for the $c$-space to~6 while 
the maximum polynomial degree for the $\varphi$-space is~4. The meshes are not
that different in the beginning of the calculation. However, one can also see that the mesh 
refinement for $c$ at $t=3.0$~s is notably different compared to $\varphi$. For instance,
the highest polynomial degree for $c$-space is~8 whereas for $\varphi$-space is~4.
Since these results are representative for all adaptivity modes, only multi-mesh 
configurations are considered in the following.
